---
title: "Occurrence Records"
format: html
---

  - `rgbif` to download records
  -  Simple filtering/removing duplicates
  -  Extract features at each point
  
## Using `rgbif` to get occurence records

The package `rgbif` enables you to access GBIF's API and request records downloads. Download requests are associated with a DOI, making them reproducible and citeable.
  
### Set up GBIF credentials

To download from GBIF, you need to register and supply your registration credentials.

More on this [here](https://docs.ropensci.org/rgbif/articles/gbif_credentials.html)

```{r}
#| eval: false
# install.packages("usethis")

usethis::edit_r_environ()

```

This will bring up a file called **.Renviron**. Add your credential information here (replace with your info).

```{r}
#| eval: false

GBIF_USER="fakeuser"
GBIF_PWD="safe_fake_password_123"
GBIF_EMAIL="fakeuser@gbif.org"

```

### Set up an `rgbif` query

There are many query parameters available to narrow down a request for observation records. We are including a few parameters that indicate acceptable-quality data for creating robust distribution maps. These include:

- Has geographic coordinates
- Has no geospatial issues
- Has acceptable geospatial coordinate accuracy
- Represents the presence of the target organism
- Within our desired time range
- Represents a human observation

